---
title: "Maths for Computing - Lecture 4"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
```


## Introduction

The objectives for today are:

 - Learn to compute determinants;
 
 - Define the concept of inverse of a matrix;
 
 - Identify the minors and the rank of a matrix;
 
 - Determine the eigenvalues and eigenvectors of a matrix;
 
 - Discuss the concept of matrix factorizations.
 
## The Determinant

![](https://youtu.be/kNr2jDWjyIs)

Let $\boldsymbol{A}$ be a $n\times n$ matrix.

The expansion of $|\boldsymbol{A}|$ in terms of the elements of the ith row is given by
$$
|\boldsymbol{A}|=a_{i,1}C_{i,1}+a_{i,2}C_{i,2}+\cdots + a_{i,n}C_{i,n},
$$
where $C_{i,j}$ is a *cofactor*.


A cofactor $C_{i,j}$ can be found as follows:

 - Delete the i-th row and the j-th column from $\boldsymbol{A}$ and compute its determinant

 - Multiply the determinant by the factor $(-1)^{i+j}$

### An example

Consider the matrix 
$$
\boldsymbol{A}=\left(
\begin{matrix}
3 & 0 & 0 &2 \\
6 & 1 & c &2 \\
-1 & 1 & 0 & 0 \\
5 & 2 & 0 & 3
\end{matrix}
\right)
$$

 - What is $C_{2,3}$? 

$C_{2,3}$ is the determinant of the submatrix of $\boldsymbol{A}$ obtained by deleting the second row and the third column multiplied by $(-1)^{2+3}=-1$. So we need to compute

$$
\det\left(
\begin{matrix}
3 & 0 & 2\\
-1 & 1 & 0\\
5 & 2 & 3
\end{matrix}
\right)
$$

To compute this determinant we can consider an expansion over any row or column. A convenient choice is one where there is one zero (for instance the first row or the third column). Let's pick the first row

\begin{align*}
\det\left(
\begin{matrix}
3 & 0 & 2\\
-1 & 1 & 0\\
5 & 2 & 3
\end{matrix}
\right) & = 3\cdot (-1)^{1+1}\det\left(\begin{matrix}1 &0\\ 2 & 3\end{matrix}\right) + 0\cdot (-1)^{1+2}\det\left(\begin{matrix} -1 &0\\ 5 & 3\end{matrix}\right) + 2\cdot (-1)^{1+3}\det\left(\begin{matrix}-1 &1\\ 5 & 2\end{matrix}\right)\\
&= 3(3 - 0) + 2(-2 -5)= 9 - 14 = -5
\end{align*}

Thus $C_{2,3}=(-1)\cdot(-5)=5$.

 - What is $|\boldsymbol{A}|$?
 
We can take advantage of having already computed $C_{2,3}$. Furthermore, although we can compute the determinant by expanding over any row or column, choosing the third column is a particularly smart choice since this has only one non-zero entry. Indeed
$$
|\boldsymbol{A}|= a_{1,3}C_{1,3} + a_{2,3}C_{2,3} + a_{3,3}C_{3,3} + a_{4,3}C_{4,4}
$$
but $a_{1,3}=a_{3,3}=a_{4,3}=0$ and therefore $|\boldsymbol{A}|=5c$.

### A Review Quiz

Consider the matrix 
$$
\boldsymbol{A}=\left(\begin{matrix}
2 & 1 & 3\\
3 & 2 & 1 \\
0 & 0 & 4
\end{matrix}
\right)
$$
```{r quiz1, echo = FALSE}
quiz(
  question(
    "What is $C_{3,3}$?",
    answer("-1"),
    answer("1",correct = TRUE),
    answer("2"),
    answer("-2"),
    allow_retry = TRUE
  ),
  question(
    "What is $|\\boldsymbol{A}|$?",
    answer("4",correct = TRUE),
    answer("-4"),
    answer("1"),
    answer("-1"),
    allow_retry = TRUE
  )
)
```

## Properties of the Determinant

Let $\boldsymbol{A}$ and $\boldsymbol{B}$ be  $n\times n$ matrices and $\alpha\in\mathbb{R}$.

 - If all elements in a row (or column) of $\boldsymbol{A}$ are 0, then $|\boldsymbol{A}|=0$

 - $|\boldsymbol{A}|=|\boldsymbol{A}^t|$: the determinant of a matrix is equal to the determinant of its transpose;

 - $|\boldsymbol{A}\boldsymbol{B}|=|\boldsymbol{A}||\boldsymbol{B}|$: the determinant of a product is the product of the determinants;

 - $|\alpha\boldsymbol{A}|=\alpha^n|\boldsymbol{A}|$

 - $|\boldsymbol{I}|=1$: the determinant of the identity matrix is one;

 - If $\boldsymbol{A}$ is triangular (or diagonal) $|\boldsymbol{A}|=\prod_{i=1}^na_{i,i}$: the determinant of a triangular matrix is equal to the product of the elements on the diagonal;

### An example
Compute the determinant of the following matrix
$$
\boldsymbol{A}=\left(
\begin{matrix}
a_1-x& a_2 & a_3 & a_4\\
0 & -x & 0 & 0 \\
0 & 1 & -x &0 \\
0 & 3 & 1 & -x
\end{matrix}
\right)
$$
Computing this determinant can be challenging if we do not pick the right expansion, but, if we do, then it becomes very simple. A smart choice is to pick the first column so that 
$$
|\boldsymbol{A}|= (a_1-x)(-1)^{1+1}\det\left(\begin{matrix}-x&0&0\\1&-x&0\\ 3 & 1 & -x\end{matrix}\right)
$$
Notice that next we need to compute the determinant of a lower triangular matrix, which is simply equal to the product of the diagonal entries. So
$$
|\boldsymbol{A}|=(a_1-x)(-x)^3
$$

### A Review Quiz

Try using the properties of determinants when answering the questions below.

```{r quiz2, echo = FALSE}
quiz(
  question(
    "What is the determinant of $\\left(\\begin{matrix} 1 & 0 &0 \\\\ 2 & 2 & 1 \\\\ 3 & 0 & 4 \\end{matrix}\\right)$?",
    answer("0"),
    answer("2"),
    answer("6"),
    answer("8",correct = TRUE),
    allow_retry = TRUE
  ),
  question(
    "What is the determinant of $\\left(\\begin{matrix} 1 & 2 &1 \\\\ 0 & 0 & 3 \\\\ 0 & 0 & 1 \\end{matrix}\\right)$?",
    answer("0",correct = TRUE),
    answer("2"),
    answer("6"),
    answer("8"),
    allow_retry = TRUE
  ),
    question(
    "What is the determinant of $\\left(\\begin{matrix} 1 & 0 \\\\ 0 & 2  \\end{matrix}\\right)\\left(\\begin{matrix} 2 & 1 \\\\ 1 & 2  \\end{matrix}\\right)$?",
    answer("0"),
    answer("2"),
    answer("6",correct = TRUE),
    answer("8"),
    allow_retry = TRUE
  )
)
```

## Matrix Inverse

![](https://youtu.be/oYvY20Q3-WI)

Given a matrix $\boldsymbol{A}$, we say that $\boldsymbol{A}^{-1}$ is its inverse if 

$$
\boldsymbol{A}^{-1}\boldsymbol{A}=\boldsymbol{A}\boldsymbol{A}^{-1}=\boldsymbol{I}
$$

Then $\boldsymbol{A}$ is said to be invertible. Notice that only square matrices can have inverses.  However, not all square matrices have an inverse. The inverse exists if and only if $det(\boldsymbol{A}) \neq 0$. If an inverse exists, this is unique.

### An Exercise
Show that $\boldsymbol{A}$ and $\boldsymbol{X}$ are inverse of each other.
$$
\boldsymbol{A}=\left(
\begin{matrix}
5 & 6 \\
5 & 10
\end{matrix}
\right)\;\;\;\;
\boldsymbol{X}=\left(
\begin{matrix}
1/2 & -3/10\\
-1/4 & 1/4
\end{matrix}
\right)
$$

The way to solve this is by either computing $\boldsymbol{A}\boldsymbol{X}$ or $\boldsymbol{X}\boldsymbol{A}$ and noticing that it equals $\boldsymbol{I}$.
$$
\boldsymbol{A}\boldsymbol{X}=\left(
\begin{matrix}
5 & 6 \\
5 & 10
\end{matrix}
\right)\left(
\begin{matrix}
1/2 & -3/10\\
-1/4 & 1/4
\end{matrix}
\right)= \left(
\begin{matrix}
5\cdot 1/2 - 6\cdot 1/4 & - 5\cdot 3/10 + 6\cdot 1/4 \\
5\cdot 1/2 - 10\cdot 1/4 & -5\cdot 3/10 + 10\cdot 1/4
\end{matrix}
\right)=
\left(
\begin{matrix}
1 & 0\\
0 & 1
\end{matrix}
\right)=
$$

### Properties of the Inverse 

Let $\boldsymbol{A}$ and $\boldsymbol{B}$ be invertible $n\times n$ matrices. Then

 - $\boldsymbol{A}^{-1}$ is invertible and $(\boldsymbol{A}^{-1})^{-1}=\boldsymbol{A}$: the inverse of the inverse of a matrix is the matrix itself;

 -  $\boldsymbol{A}\boldsymbol{B}$ is invertible and $(\boldsymbol{A}\boldsymbol{B})^{-1}=\boldsymbol{B}^{-1}\boldsymbol{A}^{-1}$: the inverse of a product is the product of the inverses in reverse order;

 - The transpose $\boldsymbol{A}^t$ is invertible and $(\boldsymbol{A}^t)^{-1}=(\boldsymbol{A}^{-1})^t$: the inverse of a transpose is the transpose of the inverse;
 
- $(c\boldsymbol{A})^{-1}=c^{-1}\boldsymbol{A}^{-1}$ for $c\neq 0$.

### Computing the inverse of a $2\times 2$ matrix

Provided that $|\boldsymbol{A}|\neq 0$,
$$
\boldsymbol{A}=\left(
\begin{matrix}
a & b\\
c& d
\end{matrix}
\right)
\longrightarrow
\boldsymbol{A}^{-1}=\frac{1}{ad-bc}\left(
\begin{matrix}
d& -b\\
-c & a
\end{matrix}
\right)
$$

### Computing the inverse of a matrix

Any $n\times n$ square matrix $\boldsymbol{A}$ has an inverse $\boldsymbol{A}^{-1}$ given by
$$
\boldsymbol{A}^{-1}=\frac{1}{|\boldsymbol{A}|}\mbox{adjoint}(\boldsymbol{A}),
$$
with 
$$
\mbox{adjoint}(\boldsymbol{A})=\left(
\begin{matrix}
C_{1,1} & C_{2,1} & \cdots & C_{n,1}\\
C_{1,2} & C_{2,2} & \cdots & C_{n,2}\\
\vdots & \vdots & \vdots & \vdots\\
C_{1,n} & C_{2,n}& \cdots & C_{n,n}
\end{matrix}
\right)
$$
where $C_{i,j}$ is the $(i,j)$ cofactor.


```{r quiz3, echo = FALSE}
quiz(
  question(
    "Which of the following matrices are invertible?",
    answer("$\\left(\\begin{matrix} 1 & 0 \\\\ 0 & 1  \\end{matrix}\\right)$",correct = TRUE),
    answer("$\\left(\\begin{matrix} 1 & 0 \\\\ 2 & 0  \\end{matrix}\\right)$"),
    answer("$\\left(\\begin{matrix} 0 & 0 \\\\ 0 & 0  \\end{matrix}\\right)$"),
    answer("$\\left(\\begin{matrix} 1 & 2 \\\\ -1 & 2  \\end{matrix}\\right)$",correct = TRUE),
    allow_retry = TRUE
  ),
 question(
    "What's the inverse of $\\boldsymbol{I}$?",
    answer("$\\left(\\begin{matrix} 1 & 0 \\\\ 0 & 1  \\end{matrix}\\right)$",correct = TRUE),
    answer("$\\left(\\begin{matrix} 1 & 0 \\\\ 2 & 0  \\end{matrix}\\right)$"),
    answer("$\\left(\\begin{matrix} 0 & 0 \\\\ 0 & 0  \\end{matrix}\\right)$"),
    answer("$\\left(\\begin{matrix} 1 & 2 \\\\ -1 & 2  \\end{matrix}\\right)$"),
    allow_retry = TRUE
 )
)
```

## Minors and Rank

### Submatrices 

A submatrix of $\boldsymbol{A}$ is a matrix obtained by deleting some rows and/or some columns.

Consider for instance the matrix:
$$ 
\left(\begin{matrix}
12 & 1 & -7 & 0.5 \\
\frac{3}{4} & -4 & 8 & 1\\
	0 & -2 & \frac{-2}{7} & 9
		\end{matrix}\right)
$$
Then 

 - $(12)$ is the submatrix obtained by deleting the second and third row and the second, the third and fourth column.
 
 - $(-4, 1)$ is the submatrix obtained by deleting the first and the third row and the first and third column.

### Minors 

A minor of a matrix $\boldsymbol{A}$ is the determinant of a square submatrix of $\boldsymbol{A}$. The minor is of *order* $k$ if it is the determinant of a submatrix $k\times k$.

Consider for example:

$$ 
A= \left( \begin{matrix}
		1 & 3 & 0 \\
		4 & 2 & 1 \\
		5 & 0 & 4 
	\end{matrix} \right)
$$

 - $\det(1)=1$ is a minor of order $1$, as well as $3,0,4,2,\ldots$

 - $\det\left( \begin{matrix} 1 & 3 \\ 4 & 2 \end{matrix}  \right)= -10$ is a minor of order $2$

 -  $\det\left( \begin{matrix} 4 & 1 \\ 5 & 4 \end{matrix}  \right)= 11$ is a minor of order 2. 

- $\det(\boldsymbol{A})$ is the only minor of order 3.

### Rank

The rank of a matrix is the order of the largest non-zero minor. The rank of $\boldsymbol{A}$ is denoted as $rank(\boldsymbol{A})$.


- $1 \leq rank(A) \leq \max\{m,n\}$ for all $m\times n$ matrices different from $\boldsymbol{0}$

- Let $\boldsymbol{A}$ be a $n\times n$ matrix. If $|\boldsymbol{A}|\neq 0$ then $rank(\boldsymbol{A})=n$. Furthermore, the matrix $\boldsymbol{A}$ is invertible if and only if it has rank $n$.
	
```{r quiz4, echo = FALSE}
quiz(
  question("What is the rank of $\\boldsymbol{I}$ of size 2?",
           answer("0"),
           answer("1"),
           answer("2",correct = TRUE),
           answer("3"),
           allow_retry=TRUE
             ),
  question("What is the rank of a $2\\times 2$ matrix with all zeros?",
           answer("0",correct = TRUE),
           answer("1"),
           answer("2"),
           answer("3"),
           allow_retry=TRUE),
  question("What is the rank of $\\left(\\begin{matrix} 1 & 1 \\\\ 0 & 0 \\end{matrix}\\right)$?",
           answer("0"),
           answer("1",correct = TRUE),
           answer("2"),
           answer("3"),
           allow_retry=TRUE)
)
```

## Eigenvalues and Eigenvectors

A different set of numbers that summarize properties of a squared matrix are the so-called *eigenvalues*. These are used widely in many real-world applications, including 

 - Google PageRank algorithms
 
 - PCA dimensionality reduction
 
 - Probabilistic models

The vector $\boldsymbol{v}$ is an eigenvector of the $n\times n$ matrix $\boldsymbol{A}$ if
		
$$ 
\boldsymbol{A}\boldsymbol{v} = \lambda \boldsymbol{v}
$$ 
		where $\lambda \in \mathbb{R}$ is an eigenvalue. 
		
### Computing eigenvalues

The eigenvalues of $\boldsymbol{A}$ are the solution to the equation
$$
p(\lambda)=det(\boldsymbol{A}-\lambda\boldsymbol{I})=0,
$$

where $p(\lambda)$ is the *characteristic polynomial*.

Let's look at an example. Consider the matrix

$$
\boldsymbol{A} = \left(\begin{matrix} -1 &  2 \\ 3 &0   \end{matrix}  \right) 
$$
We need to solve the equation

$$
\det(\boldsymbol{A}-\lambda \boldsymbol{I}) = 
		\det\left( \begin{matrix} 
			-1 -\lambda & 2 \\ 3 & -\lambda  
		\end{matrix}\right) = 0 
$$
which, using the rules for determinants, is
$$
\lambda^2 + \lambda - 6 = 0  
$$
and therefore
$$   
\lambda_1 = -3 \quad \lambda_2 = 2 
$$  
		
### Computing eigenvectors

Once an eigenvalue $\lambda$ is found, it is possible to find the associated eigenvectors by solving
$$
\boldsymbol{A}\boldsymbol{v} = \lambda \boldsymbol{v} 
$$

Notice that there is an infinite number of eigenvectors associated to each eigenvalue (this will become apparent below).

Let's find the eigenvectors associated to the eigenevalue $2$ of the previous example. We have to solve
$$
\left(\begin{matrix} -1 &  2 \\ 3 &0   \end{matrix}  \right)
			\left(\begin{matrix} v_1 \\ v_2 \end{matrix}\right) =
			   2 \left(\begin{matrix} v_1 \\ v_2 \end{matrix}\right)
$$
which is a linear system of two equations with two unknowns,
$$
\left\{ \begin{matrix} -v_1 + 2 v_2 = 2v_1 \\
		3v_1 = 2v_2 \end{matrix} \right.
$$  
transformed to,
$$
\left\{ \begin{matrix} -3v_1 + 2 v_2 = 0  \\
		3v_1  -  2v_2 = 0 \end{matrix} \right.
$$  
The two equations are equal and therefore we are left with one only
$$
3v_1  -  2v_2 = 0
$$
whose solutions different from zero are,
$$
v_2 = \frac{3}{2} v_1 
$$
So \textbf{one} solution is  
$$
v_1 = 2 , v_2 = 3 
$$
Let's check,
$$
\left(\begin{matrix} -1 &  2 \\ 3 &0   \end{matrix}  \right)
			\left(\begin{matrix} 2 \\ 3 \end{matrix}\right)=
		\left(\begin{matrix} -2 + 6 \\ 6  \end{matrix}\right)=
			\left(\begin{matrix} 4 \\ 6 \end{matrix}\right)=
			2\left(\begin{matrix} 2 \\ 3 \end{matrix}\right)
$$

All possible eigenvectors (with eigenvalue 2) can be written as
$$
\left(\begin{matrix} 2t \\ 3t \end{matrix}\right) 
			\quad t \in \mathbb{R}
$$
So the following are also eigenvectors
$$
(-2, -3), (4,6) , (2\pi, 3\pi), (-20,-30), ...
$$


```{r quiz5, echo=FALSE}
quiz(
  question("Which of the following is an eigenvalue of $\\boldsymbol{I}$ of size 2?",
         answer("0"),
         answer("1",correct=TRUE),
         answer("2"),
         answer("3"),
         allow_retry = TRUE),
question("Which of the following are eigenvectors of $\\boldsymbol{I}$ of size 2?",
         answer("(0,1)",correct = TRUE),
         answer("(1,0)",correct = TRUE),
         answer("(2,2)",correct= TRUE),
         answer("(1,2)",correct = TRUE),
         allow_retry = TRUE
)
)
         
```

## Matrix Factorizations

A factorization of a matrix $\boldsymbol{A}$ is a way to write $\boldsymbol{A}$ as a product of matrics, simpler or with some specific characteristics

 - Useful to compute the determinant 
 
 - To compute the matrix inverse $\boldsymbol{A}^{-1}$ 

 - To solve linear systems of equations

 - For any other reason
 
Basically all functions implemented in software to compute the above quantities implicitly exploits such factorizations to speed up computations.

### LU Factorization

If $det(\boldsymbol{A}) \neq 0$ then,	$$ \boldsymbol{A} = \boldsymbol{L}\boldsymbol{U} $$ 
	where $\boldsymbol{A},\boldsymbol{L},\boldsymbol{U}$ are $n\times n$ matrices, $\boldsymbol{L}$ is lower triangular and 
	$\boldsymbol{U}$ is upper triangular.
	
 - The LU factorization is valid for squared matrices only. It writes a squared matrix as the product of a lower triangular and an upper triangular matrix.
 
 - The matrices $\boldsymbol{L},\boldsymbol{U}$ are unique for every $\boldsymbol{A}$   

 - If $ \boldsymbol{A} = \boldsymbol{L}\boldsymbol{U} $ then it is very easy to compute the determinant. This is because the determinant of a product is the product of the determinants and the determinant of a triangular matrix is just the product of its diagonal entries.
 
### Cholesky Factorization

If $\boldsymbol{A} = \boldsymbol{A}^t$ and $\boldsymbol{A}$ is positive-definite,then
	 $$ \boldsymbol{A} = \boldsymbol{L}\boldsymbol{L}^t$$
	 where $\boldsymbol{L}$ is lower triangular.
	 
 - Cholesky factorization only applies to symmetric (thus squared) matrices which have the additional property of being positive-semidefinite.
 
 - A symmetric matrix $\boldsymbol{A}$ is said to be positive-definite if for all non-zero column vectors $\boldsymbol{x}$, $\boldsymbol{x}^t\boldsymbol{A}\boldsymbol{x}>0$.
 
 - The property of positive-semidefiniteness and Cholesky factorizations are widely used in statistics.

### Matrix Diagonalization

A $n\times n$ matrix $\boldsymbol{A}$ is diagonalizable if there exist an invertible $n\times n$ matrix $\boldsymbol{P}$ and a $n\times n$ diagonal matrix	 $\boldsymbol{D}$ such that
		$$  \boldsymbol{A} = \boldsymbol{P}\boldsymbol{D}\boldsymbol{P}^{-1} $$

 - The elements on the diagonal of $\boldsymbol{D}$ are the eigenvalues

 - The columns of $\boldsymbol{P}$ are the eigenvectors
 
 - Not all matrices are diagonalizable 
 
 - If $\boldsymbol{A}$ is symmetric then it is diagonalizable 

	
